Evaluating Sound Similarity Metrics for Differentiable, Iterative Sound-Matching

### 2025-09-24 sections/background_related.tex
Updates from Overleaf

Having looked at the current literature, we identify four major areas of weakness in sound-matching. We target the first two issues in this work, while the latter two issues are left for future work.
\begin{enumerate}
    \item \LossSelect: It remains unclear whether there exists a universally best-performing loss function, 
    or whether performance depends on factors such as the sound domain, synthesizer architecture, 
    and desired output characteristics. Existing studies typically evaluate losses only with simple synthesis 
    setups~\cite{vahidi2023mesostructures}, leaving open questions about their generality. 
    \item \SynthSelect: There is a lack of diversity in the synthesis methods used in sound-matching. The effects of using different isolated DSP functions in iterative sound-matching and how these interact with loss functions remains largely untested.
    \item \PeriodicLoss: The sinusoidal nature of loss function landscapes is a problem that appears frequently\todo{cite and make clearer}. In differentiable settings, this problem causes gradient descent updates to local minima. 
    \item \OutDomain: Sound-matching with out-of-domain sounds is a largely unexplored area, yet a necessary one for practical applications for sound designers.
\end{enumerate}

>>>>

Having looked at the current literature, we identify four major areas of weakness in sound-matching. We target the first two issues in this work, while the latter two issues are left for future work.
\begin{enumerate}
    \item \LossSelect: It remains unclear whether there exists a universally best-performing loss function, 
    or whether performance depends on factors such as the sound domain, synthesizer architecture, 
    and desired output characteristics. Existing studies typically evaluate losses only with simple synthesis 
    setups~\cite{vahidi2023mesostructures}, leaving open questions about their generality. 
    \item \SynthSelect: There is a lack of diversity in the synthesis methods used in sound-matching. The effects of using different isolated DSP functions in iterative sound-matching and how these interact with loss functions remains largely untested.
    \item \PeriodicLoss: Loss function landscapes are not easy to navigate, as there can be many local minima, or large flat areas~\cite{turian2020sorry,vahidi2023mesostructures}.  In differentiable settings, this can cause gradient descent updates to not reach the global minima. 
    \item \OutDomain: Sound-matching with out-of-domain sounds is a largely unexplored area, yet a necessary one for practical applications for sound designers.
\end{enumerate}

### 2025-09-24 sections/summary_weakness_conclusion.tex
Updates from Overleaf

\subsection{Key Findings}
The most important takeaway regarding our main hypothesis is that the loss function that yields the best sound-matching outcomes varies depending on the synthesizer program. This is likely due to the interaction between how the parameters of the synthesizer influence the sound, and the core sonic features used by the loss function.

@@

>>>>

\section{Discussion}
\subsection{Key Findings}
The most important takeaway regarding our main hypothesis is that the loss function that yields the best sound-matching outcomes varies depending on the synthesizer program. This is likely due to the interaction between how the parameters of the synthesizer influence the sound, and the core sonic features used by the loss function.

@@

\subsection{Illustrating Loss Landscapes}
\label{sec:loss_landscape_examples}
To complement our quantitative analysis, we provide examples of loss landscapes that illustrate why specific losses succeed or fail in certain synthesis contexts. These are illustrative aids rather than conclusive evidence, and highlight the importance of \PeriodicLoss{} discussed earlier in Section~\ref{sec:lacking}.

We show the landscapes of the loss functions with simplified versions of \BPNoise{} (HP-Noise) and \AmpMod{} (S-Noise-AM), each reduced to a single parameter for one-dimensional visualization. HP-Noise uses only a high-pass filter with cutoff between 100–20,000 Hz, while S-Noise-AM varies only the modulation rate between 0.1–20 Hz. Loss values are normalized to 0–1 to allow comparison across functions.

Figure~\ref{fig:loss_landscape_noisebp} shows the HP-Noise results. \LoneSpec{}, \SIMSESpec{}, and \JTFS{} exhibit clear minima at the correct parameter (red dashed line), whereas \DTWEnv{} remains relatively flat, making gradient-based optimization more difficult.

Figure~\ref{fig:loss_landscape_ampmod} shows the S-Noise-AM results. Here, \DTWEnv{} produces the smoothest and most informative landscape, while spectrogram losses are minimized near the target but lack consistent correlation with parameter distance. \JTFS{} remains largely flat. This aligns with \DTWEnv's superior performance in amplitude-modulated synthesis observed in our experiments.

\begin{figure*}[ht]
    \centering
    \begin{minipage}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\linewidth]{images/experiment_plots/comparing_loss_landscapes_normalized.png}
        \caption{Loss landscapes for \BPNoise{} with only a high-pass filter parameter. 
        \LoneSpec{}, \SIMSESpec{}, and \JTFS{} show clear global minima near the correct parameter, while \DTWEnv{} remains flat around the target.}
        \label{fig:loss_landscape_noisebp}
    \end{minipage}%
    \hfill
    \begin{minipage}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\linewidth]{images/experiment_plots/comparing_lanscapes_1_1d_normalized.png}
        \caption{Loss landscapes for a simplified \AmpMod{} synthesizer with only amplitude modulation parameter. 
        \DTWEnv{} exhibits the smoothest and most informative landscape, explaining its superior performance in amplitude-modulated synthesis.}
        \label{fig:loss_landscape_ampmod}
    \end{minipage}
\end{figure*}
